---
title: "Data Science Applied to Ag - Final Project - ML"
format:
  html:
    embed-resources: true
    toc: true
    theme: cerulean
author: Md Shakir Moazzem, Umar Munir
---

# Introduction  

This script contains the data wrangling and feature engineering steps for the final project.

# Start of data wrangling

## Loading packages  

The following code chunk will load necessary packages.

```{r Setup, message=F}

# Installing packages

#install.packages("tidyverse")
#install.packages("readxl")
#install.packages("janitor")
#install.packages("dplyr")
#install.packages("tidyr")
#install.packages("readr")
#install.packages("lubridata")
#install.packages("stringr")

# Loading packages 

library(tidyverse)
library(readxl) # to read excel files
library(janitor) # to clean data; helps fix and standardize the column names
library(dplyr) # wrangling
library(tidyr) # wrangling
library(readr) # to export csv
library(lubridate)
library(stringr)

```

## Reading "training" data  

The following code chunk will read the csv files for the 3 training data sets

```{r training data import, message=F, warning=F}

#reading the csv files for the 3 training data sets 

trait_df <- read_csv("../data/training/training_trait.csv") 
meta_df  <- read_csv("../data/training/training_meta.csv")
soil_df  <- read_csv("../data/training/training_soil.csv")

submission_test <- read_csv("../data/testing/testing_submission.csv")
soil_test <- read_csv("../data/testing/testing_soil.csv")
meta_test <- read_csv("../data/testing/testing_meta.csv")


```


## Data wrangling on "training" data

The code below creates a function to adjust the yield_mg_ha for 15.5% grain moisture.

```{r}

# Function to transform yield to 15.5% moisture
adjust_yield <- function(yield_mg_ha, grain_moisture) {
  yield_mg_ha * (100 - grain_moisture) / (100 - 15.5)
}

```


